---
title: "PSTAT 196 Week 1 Write-Up"
author: "Terry Jiang, John Randazzo, Jason Freeberg"
date: "4/17/2017"
output: word_document
---

```{r, include = FALSE}
transitioned <- read.csv("196 transition.csv")
censored <- read.csv("196 censor.csv")
full <- rbind(transitioned,censored)
full$event <- as.numeric(full$Partial_code_ff)
library(ggplot2)
library(survival)
library(plyr)
library(dplyr)
```

##Exploring Our Data

- We are using the following covariates in our analysis:
sex, Socio.Economic, alcohol, cigar, BMInew.

- Our time variable is DURATION

- For each dataset, we define a binary event variable, event = as.numeric(Partial_code_ff)

- Our transitioned dataset is 113167x17

- Our censored dataset is 13498x17

- Our full dataset is therefore 126665x17

- We also need to take care of the occasional duplicate measure for Socio.Economic reclassification.


## Further Summary Statistics

```{r, echo = FALSE}
table.sex <- rbind(table(transitioned$sex),table(censored$sex),table(full$sex)) #Summarizing sex counts
rownames(table.sex) <- c("Trasnsitioned","Censored","All")
colnames(table.sex) <- c("Male","Female")
```

Number of men and women in the study, and then their relative proportions:

```{r, echo = FALSE}
table.sex
prop.table(table.sex,1) #Show row percentages of sex
```

Socio-economic classification counts for each dataset:

```{r, echo = FALSE}
table.socio <- rbind(table(transitioned$Socio.Economic),
                     table(censored$Socio.Economic),
                     table(full$Socio.Economic)) #Summarizing socio.economic status counts
table.socio <- as.data.frame.matrix(table.socio) #Convert the table to a data frame
table.socio$mean <- c(mean(transitioned$Socio.Economic),
                      mean(censored$Socio.Economic),
                      mean(full$Socio.Economic)) #Add a mean colume to the data frame
rownames(table.socio) <- c("Trasnsitioned","Censored","All")
table.socio
```

Summary statistics for BMI - These are a bit wonky. We will revisit this table a bit later in the report.
```{r, echo = FALSE}

table.BMI <- rbind(summary(transitioned$BMInew),
                   summary(censored$BMInew),
                   summary(full$BMInew)) #Summarizing the BMI
rownames(table.BMI) <- c("Trasnsitioned","Censored","All")
table.BMI
```

Percentages of smokers and drinkers in each dataset:

```{r, echo = FALSE}

table.cigalc <- matrix(c(mean(transitioned$cigar),mean(transitioned$alcohol),
                         mean(censored$cigar),mean(censored$alcohol),
                         mean(full$cigar),mean(full$alcohol)),ncol=2,byrow=TRUE) #Summarizing smoker and alcohol user percentage
rownames(table.cigalc) <- c("Trasnsitioned","Censored","All")
colnames(table.cigalc) <- c("Smoker","Used Alcohol")
table.cigalc
```

```{r, echo = FALSE}

table.duration <- rbind(summary(transitioned$DURATION),
                   summary(censored$DURATION),
                   summary(full$DURATION)) #Summarizing the duration
rownames(table.duration) <- c("Trasnsitioned","Censored","All")
```

Summary statistics for DURATION:

```{r, echo = FALSE}
table.duration
```

## A Few Words About BMI and Other Data Management Issues

We found some funky BMI values for certain subjects in the study.

```{r}
max(full$BMInew)
min(full$BMInew)
boxplot(full$BMInew, ylab = "BMInew")
```

The population for our study consists of people with an existing history of moderate hypertension; a health problem generally associated with an unhealthy body weight. However, we believe that a BMI of 0.2 or 66000.0 is likely the product of clerical data entry error, and inspecting the boxplot confirms that there are many other outlandishly incorrect BMI measures in our dataset. We need to remove these broken values:

```{r}
full2 <- full
full2$BMInew <- ifelse(full2$BMInew > 100 | full2$BMInew < 12, NA, full2$BMInew)
full2 <- na.omit(full2)
```

After considering the origins of our data and the range of values that a person with moderate hypertension's BMI can take, we decide to omit any BMI measures under 12.00 or above 100.00.

With junk values removed, our new dataset, full2, is now 126339x17.

The new table of summary statistics for each dataset's BMI measures is as such:

```{r, echo = FALSE}
censored2 <- censored
censored2$BMInew <- ifelse(censored2$BMInew > 100 | censored2$BMInew < 12, NA, censored2$BMInew)
censored2 <- na.omit(censored2)

transitioned2 <- transitioned
transitioned2$BMInew <- ifelse(transitioned2$BMInew > 100 | transitioned2$BMInew < 12, NA, transitioned2$BMInew)
transitioned2 <- na.omit(transitioned2)

table.BMI2 <- rbind(summary(transitioned2$BMInew),
                   summary(censored2$BMInew),
                   summary(full2$BMInew)) #Summarizing the BMI
rownames(table.BMI2) <- c("Trasnsitioned","Censored","All")
table.BMI2
```


## Exploratory Kaplan-Meier Estimates

Now that we have a good feel for our dataset, we are ready and eager to begin our survival analysis. We compute Kaplan Meier estimates contingent on each covariate's different levels, as well as a general KM estimate for our entire population. 

```{r, echo = FALSE}

# From: https://www.r-statistics.com/2013/07/creating-good-looking-survival-curves-the-ggsurv-function/
ggsurv <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                   cens.col = 'red', lty.est = 1, lty.ci = 2,
                   cens.shape = 3, back.white = F, xlab = 'Time',
                   ylab = 'Survival', main = ''){
  
  library(ggplot2)
  strata <- ifelse(is.null(s$strata) ==T, 1, length(s$strata))
  stopifnot(length(surv.col) == 1 | length(surv.col) == strata)
  stopifnot(length(lty.est) == 1 | length(lty.est) == strata)
  
  ggsurv.s <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                       cens.col = 'red', lty.est = 1, lty.ci = 2,
                       cens.shape = 3, back.white = F, xlab = 'Time',
                       ylab = 'Survival', main = ''){
    
    dat <- data.frame(time = c(0, s$time),
                      surv = c(1, s$surv),
                      up = c(1, s$upper),
                      low = c(1, s$lower),
                      cens = c(0, s$n.censor))
    dat.cens <- subset(dat, cens != 0)
    
    col <- ifelse(surv.col == 'gg.def', 'black', surv.col)
    
    pl <- ggplot(dat, aes(x = time, y = surv)) +
      xlab(xlab) + ylab(ylab) + ggtitle(main) +
      geom_step(col = col, lty = lty.est)
    
    pl <- if(CI == T | CI == 'def') {
      pl + geom_step(aes(y = up), color = col, lty = lty.ci) +
        geom_step(aes(y = low), color = col, lty = lty.ci)
    } else (pl)
    
    pl <- if(plot.cens == T & length(dat.cens) > 0){
      pl + geom_point(data = dat.cens, aes(y = surv), shape = cens.shape,
                      col = cens.col)
    } else if (plot.cens == T & length(dat.cens) == 0){
      stop ('There are no censored observations')
    } else(pl)
    
    pl <- if(back.white == T) {pl + theme_bw()
    } else (pl)
    pl
  }
  
  ggsurv.m <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                       cens.col = 'red', lty.est = 1, lty.ci = 2,
                       cens.shape = 3, back.white = F, xlab = 'Time',
                       ylab = 'Survival', main = '') {
    n <- s$strata
    
    groups <- factor(unlist(strsplit(names
                                     (s$strata), '='))[seq(2, 2*strata, by = 2)])
    gr.name <-  unlist(strsplit(names(s$strata), '='))[1]
    gr.df <- vector('list', strata)
    ind <- vector('list', strata)
    n.ind <- c(0,n); n.ind <- cumsum(n.ind)
    for(i in 1:strata) ind[[i]] <- (n.ind[i]+1):n.ind[i+1]
    
    for(i in 1:strata){
      gr.df[[i]] <- data.frame(
        time = c(0, s$time[ ind[[i]] ]),
        surv = c(1, s$surv[ ind[[i]] ]),
        up = c(1, s$upper[ ind[[i]] ]),
        low = c(1, s$lower[ ind[[i]] ]),
        cens = c(0, s$n.censor[ ind[[i]] ]),
        group = rep(groups[i], n[i] + 1))
    }
    
    dat <- do.call(rbind, gr.df)
    dat.cens <- subset(dat, cens != 0)
    
    pl <- ggplot(dat, aes(x = time, y = surv, group = group)) +
      xlab(xlab) + ylab(ylab) + ggtitle(main) +
      geom_step(aes(col = group, lty = group))
    
    col <- if(length(surv.col == 1)){
      scale_colour_manual(name = gr.name, values = rep(surv.col, strata))
    } else{
      scale_colour_manual(name = gr.name, values = surv.col)
    }
    
    pl <- if(surv.col[1] != 'gg.def'){
      pl + col
    } else {pl + scale_colour_discrete(name = gr.name)}
    
    line <- if(length(lty.est) == 1){
      scale_linetype_manual(name = gr.name, values = rep(lty.est, strata))
    } else {scale_linetype_manual(name = gr.name, values = lty.est)}
    
    pl <- pl + line
    
    pl <- if(CI == T) {
      if(length(surv.col) > 1 && length(lty.est) > 1){
        stop('Either surv.col or lty.est should be of length 1 in order
             to plot 95% CI with multiple strata')
      }else if((length(surv.col) > 1 | surv.col == 'gg.def')[1]){
        pl + geom_step(aes(y = up, color = group), lty = lty.ci) +
          geom_step(aes(y = low, color = group), lty = lty.ci)
      } else{pl +  geom_step(aes(y = up, lty = group), col = surv.col) +
          geom_step(aes(y = low,lty = group), col = surv.col)}
    } else {pl}
    
    
    pl <- if(plot.cens == T & length(dat.cens) > 0){
      pl + geom_point(data = dat.cens, aes(y = surv), shape = cens.shape,
                      col = cens.col)
    } else if (plot.cens == T & length(dat.cens) == 0){
      stop ('There are no censored observations')
    } else(pl)
    
    pl <- if(back.white == T) {pl + theme_bw()
    } else (pl)
    pl
  }
  pl <- if(strata == 1) {ggsurv.s(s, CI , plot.cens, surv.col ,
                                  cens.col, lty.est, lty.ci,
                                  cens.shape, back.white, xlab,
                                  ylab, main)
  } else {ggsurv.m(s, CI, plot.cens, surv.col ,
                   cens.col, lty.est, lty.ci,
                   cens.shape, back.white, xlab,
                   ylab, main)}
  pl
}

density <- ggplot(data=full, mapping=aes(x=DURATION, fill=event)) +
              geom_density(alpha=0.5) +
              ggtitle("Density Plot of Surival Times") +
              xlim(0, 10000) +
              xlab("Time (days)") +
              scale_y_continuous(labels = scales::percent)

# Need to remove BMI outliers
full2 <- full
full2$BMInew <- ifelse(full2$BMInew > 100 | full2$BMInew < 15, NA, full2$BMInew)
full2 <- na.omit(full2)
bmiQuants <- quantile(full2$BMInew, seq(0, 1, 0.25))
full2$bmiCat <- cut(full2$BMInew, unique(bmiQuants), include.lowest=TRUE)
full2$bmiCat <- factor(full2$bmiCat, unique(levels(full2$bmiCat))[c(1,2,3,4)], labels=c("Low", "Medium", "High", "Obese"))

# Exploratory KM Estimates
survObject <- Surv(full$DURATION, full$event)
survObject2 <- Surv(full2$DURATION, full2$event)
bmiKM <- survfit(survObject2 ~ full2$bmiCat)
alcKM <- survfit(survObject ~ full$alcohol)
cigKM <- survfit(survObject ~ full$cigar)
seKM <- survfit(survObject ~ full$Socio.Economic)
fitKM <- survfit(survObject2 ~ 1)

ggsurv(fitKM, plot.cens=F) +
  ggtitle("Estimated Probability of Survival over Time")

ggsurv(alcKM, plot.cens=F) +
  ggtitle("KM Estimates Stratified by Alcohol Consumption") +
  scale_fill_discrete(name="New Legend Title")

ggsurv(bmiKM, plot.cens=F) +
  ggtitle("Stratified by BMI Range")

ggsurv(cigKM, plot.cens = F) +
  ggtitle("Stratified by Cigarette Usage")

ggsurv(seKM, plot.cens = F) +
  ggtitle("Stratified by Socio-Economic Classification")
```

